Liveness-Based Garbage Collection for Lazy Languages

### 2015-04-26 paper/paper.tex
Added comments for fixing liveness equation part

The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \bcar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \bcdr\sigma_2',~\mbox{and}\\  
&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
\end{align*}

@@

\section{Generating Liveness Equations}
The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
example. Given  an arbitrary function $\mathit{f}$,  we now illustrate
how  to generate  the equations  describing the  demand transformation
\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
determine   \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.    Of   particular
interest  is the liveness  at the  program point  at the  beginning of
beginning of  $e_{\mathit{f}}$ (assumed  to be $\pi_1$).   We generate
equations     of     the     form    $\Lf{\mathit{f}}{i}{\sigma}     =
\Lanv{\pi_1}{x_i}$, where  $x_i$ is the  $i^{\mbox{\footnotesize th}}$
parameter of $\mathit{f}$. Doing this for \length, we get:
\begin{align*}
 \Lanv{\pi_1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
\end{align*}
Which gives the following equation for \Lfone{\length}: 
  \begin{eqnarray*}
    \Lf{\length}{1}{\sigma}
    &=& \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
   \end{eqnarray*}

@@

As mentioned earlier, each of  the liveness rules modifies a demand by
prefixing   with   symbols   in   the   above   set,   and   therefore
$\Lf{f}{i}{\sigma}$ has the closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} &=& \Df{f}{i}.\sigma
\end{eqnarray}
where \Df{f}{i} are sets of  strings over the alphabet $\lbrace \acar,
\acdr,\bcar, \bcdr, \cbar \rbrace$.   Substituting the guessed form in
the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
$\sigma$,  we  get the  following  equation  for  \Df{f}{i}: that  are
independent of $\sigma$.  Any solution for \Df{f}{i} yields a solution
for \Lfonly$_{\mathit f}$.  Applied to \Lfonly$_{\mathit \length}$, we
get:
  \begin{align*}
    \Df{\length}{1} &= \lbrace\cbar\rbrace 
    \cup \acdr\Df{\length}{1}
  \end{align*}

Note that this equation can also be viewed as CFGs.

@@

We now calculate a 0-CFA-style summary liveness for each GC point of a
program.  There are two parts  to this.  First, for each function $f$,
we determine a {\em  summary demand} $\sigma_f$ over-approximating any
demand $\sigma$ passed to $f$.  So, consider a function $g$ containing
a  call   site  $\pi$   to  $f$,  say   $\pi\!\!:\!(\LET~x  \leftarrow
(f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.   Let   the  demand  on   $g$  be
$\sigma_g$ and the  liveness of $x$ at $\pi_i$  be $\Lanv{i}{x}$. Then
the $\LET$ case of Figure~\ref{fig:live-judge} tells us this call site
contributes  $\Lanv{i}{x}$ to  the  demand $\sigma_f$  placed on  $f$.
Write $\deltacall{f}{\pi_i}{g}$  for this demand  (contributed by call
site $\pi$  in $g$).  Thus, supposing  the $k$ call  sites to function
$f$  are  $\pi^1$ (in  function  $g^1$)  \ldots  $\pi^k$ (in  function
$g^k$),  then  the  over-approximation  requirement on  $\sigma_f$  is
achieved by  taking $\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
\cup \deltacall{f}{\pi^k}{g^k}$.

@@

For some of the program points in \length\  and \main,  this gives
{\color {red} Prasanna, please fill this correctly}. You could use the
set notation if the map notation spills over: 
\begin{align*}
\Lan{\pi_1} &= \Lan{\pi_2} = \ldots \\
\Lan{\pi_3} &= \ldots \\
\Lan{\pi_4} &= \ldots \\
\Lan{\pi_5} &= \Lan{\pi_6} = \Lan{\pi_7} \ldots \\
\Lan{\pi_8} &= \Lan{\pi_9} = \Lan{\pi_{10}} \ldots
\end{align*}
\subsection{Solving liveness equations---the grammar interpretation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{7}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{7}{\xl}}  &\rightarrow&   \var{\acdr \Lf
  {\length}{1}{\epsilon}} \mid \epsilon \\
\end{eqnarray*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Uf{f}{i} and \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
% \marginpar{Consistency: summary vs cumulative?}
\item For each function $\mathit{f}$ (including $\mainpgm$ for $e_\mainpgm$)
an equation defining
  liveness at each GC point of $e_{\mathit f}$.

\subsection{Solving liveness equations---the grammar interpretation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
% being a grammar with multiple start symbols---one for
% for each nonterminal \var{\Lanv{i}{x}}.
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{9}{y}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{9}{y}}  &\rightarrow&  \var{\Uf{\append}{1}} \mid  \var{\Df
  {\append}{1}}        (\epsilon         \mid        \acdr        \mid
 \acdr\acar\var{$\sigma_{\mathit{\!all}}$})\\
  \var{\Uf{\append}{1}}    &\rightarrow&    \epsilon   \mid
  \acdr\var{\Uf{\append}{1}}\\         
  \var{\Df{\append}{1}}
  &\rightarrow&               \acar\bcar               \mid
  \acdr\var{\Df{\append}{1}}\bcdr\\
\var{$\sigma_{\mathit{\!all}}$} &\rightarrow & \epsilon \mid \acar
\var{$\sigma_{\mathit{\!all}}$} \mid \acdr
\var{$\sigma_{\mathit{\!all}}$}
\end{eqnarray*}
Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
denoted $\mathscr{L}(\var{\Lanv{i}{x}})
$,  is the desired  solution of
\Lanv{i}{x}.   
However, recall from  our earlier  discussion that
the  decision problem that  we are  interested in  during garbage
collection is:
\begin{quote}
Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
of forward paths only. Then does $\alpha \in \sigma$?
\end{quote}
We could convert  the  rules  defining  $\hookrightarrow$  into
productions and  add them to the grammar. However, this results
in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
membership  problem   for  such  grammars   is  undecidable.   We
circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
  strongly  regular}  CFGs which  have   easy
translations to non-deterministic finite state automata (NFA). The
NFAs are then simplified on the
basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
membership of forward access paths. The resulting NFAs are   finally
converted to DFAs for use during garbage collection.
%------------------------------------------------------------%
\subsubsection{Approximating CFGs using NFAs.}
\label{sec:NFA-approx}

>>>>

The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \acdr\sigma_2',~\mbox{and}\\  
&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
\end{align*}

@@

\section{Generating Liveness Equations}
The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
example. Given  an arbitrary function $\mathit{f}$,  we now illustrate
how  to generate  the equations  describing the  demand transformation
\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
determine   \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.    Of   particular
interest  is the liveness  at the  program point  at the  beginning of
$e_{\mathit{f}}$ (assumed  to be $\pi_1$).   We generate
equations     of     the     form    $\Lf{\mathit{f}}{i}{\sigma}     =
\Lanv{\pi_1}{x_i}$, where  $x_i$ is the  $i^{\mbox{\footnotesize th}}$
parameter of $\mathit{f}$. Doing this for \length, we get:
\begin{align*}
 \Lanv{\pi_1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
\end{align*}
Which gives the following equation for \Lfone{\length}: 
  \begin{eqnarray*}
    \Lf{\length}{1}{\sigma}
    &=& \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
   \end{eqnarray*}

@@

As mentioned earlier\comment{Check if already mentioned}, each of  the liveness rules modifies a demand by
prefixing   it with   symbols   in   the   above   set,   and   therefore
$\Lf{f}{i}{\sigma}$ has the closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} &=& \Df{f}{i}.\sigma
\end{eqnarray}
where \Df{f}{i} are sets of  strings over the alphabet $\lbrace \acar,
\acdr,\bcar, \bcdr, \cbar \rbrace$.   Substituting the guessed form in
the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
$\sigma$,  we  get an  equation  for  \Df{f}{i} that is
independent of $\sigma$.  Any solution for \Df{f}{i} yields a solution
for \Lfonly$_{\mathit f}$.  Applied to \Lfonly$_{\mathit \length}$, we
get:
  \begin{align*}
    \Df{\length}{1} &= \lbrace\cbar\rbrace 
    \cup \acdr\Df{\length}{1}
  \end{align*}

Note that this equation can also be viewed as a CFG with \{\acdr,
\cbar\} as the terminal symbols and \Df{\length}{1} as the sole non-terminal.

@@

We now calculate a 0-CFA-style summary liveness for each GC point of a
program.  There are two parts  to this.  First, for each function $f$,
we determine a {\em  summary demand} $\sigma_f$ over-approximating any
demand $\sigma$ passed to $f$.  So, consider a function $g$ containing
a  call   site  $\pi$   to  $f$,  say   $\pi\!\!:\!(\LET~x  \leftarrow
(f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.   Let   the  demand  on   $g$  be
$\sigma_g$ and the  liveness of $x$ at $\pi_i$  be $\Lanv{i}{x}$. Then
the $\LET$ case of Figure~\ref{fig:live-judge} tells us this call site
contributes  $\Lanv{i}{x}$ to  the  demand $\sigma_f$  placed on  $f$.
Write $\deltacall{f}{\pi_i}{g}$  for this demand  (contributed by call
site $\pi$  in $g$).  Thus, supposing  the $k$ call  sites to function
$f$  are  $\pi^1$ (in  function  $g^1$)  \ldots  $\pi^k$ (in  function
$g^k$),  then  the  over-approximation  requirement on  $\sigma_f$  is
achieved by  taking $\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
\cup \deltacall{f}{\pi^k}{g^k}$\comment{Rewrite, copied from earlier paper}.

@@

For some of the program points in \length\  and \main,  this gives
{\color {red} Prasanna, please fill this correctly}. You could use the
set notation if the map notation spills over: 
\begin{align*}
\Lan{\pi_1} &= \Lan{\pi_2} = \ldots \\
\Lan{\pi_3} &= \ldots \\
\Lan{\pi_4} &= \ldots \\
\Lan{\pi_5} &= \Lan{\pi_6} = \Lan{\pi_7} \ldots \\
\Lan{\pi_8} &= \Lan{\pi_9} = \Lan{\pi_{10}} \ldots
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
% \marginpar{Consistency: summary vs cumulative?}
\item For each function $\mathit{f}$ (including $\mainpgm$ for
  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages?}
an equation defining
  liveness at each GC point of $e_{\mathit f}$.

\subsection{Solving liveness equations---the grammar interpretation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{7}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{7}{\xl}}  &\rightarrow&   \var{\acdr \Lf
  {\length}{1}{\epsilon}} \mid \epsilon \\
\end{eqnarray*}
\comment{Pick up a richer equation or more than one liveness equations}
Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
denoted $\mathscr{L}(\var{\Lanv{i}{x}})
$,  is the desired  solution of
\Lanv{i}{x}.   
However, recall from  our earlier  discussion that
the  decision problem that  we are  interested in  during garbage
collection is:
\begin{quote}
Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
of forward paths only. Then does $\alpha \in \sigma$?
\end{quote}
We could convert  the  rules  defining  $\hookrightarrow$  into
productions and  add them to the grammar. However, this results
in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
membership  problem   for  such  grammars   is
undecidable. \comment{Rewrite it to say that this is a CFG with 3
  unrestricted rules and the membership for even this grammar is undecidable}
\comment{State the undecidability of the decision problem above.}

  We circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
  strongly  regular}  CFGs which  have   easy
translations to non-deterministic finite state automata (NFA). The
NFAs are then simplified on the
basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
membership of forward access paths. The resulting NFAs are   finally
converted to DFAs for use during garbage collection.
\comment{Fix all comments and produce first draft}
%------------------------------------------------------------%
\subsubsection{Approximating CFGs using NFAs.}
\label{sec:NFA-approx}
